### 2014-08-14 doc/thesis/typedlua.tex
Adding text about modules

It is also not sound to allow the type of the fields to change,
as the next example shows:
\begin{verbatim}
    local person = { firstname = "Lewis", lastname = "Reed" }
    person.middlename = "Allan"
    person.firstname = 1964              -- compile-time error
    person.lastname = 2013               -- compile-time error
    print("bye bye " .. person.firstname)
\end{verbatim}

@@

Typed Lua will also introduce a specific syntax for defining modules.
For now, each file is a module.
We use all global names and their types to build the interface that
represents the type of the module, in case the module is not returned
at the end of the file.

>>>>

It is also not sound to allow the type of the fields to change,
as the next example shows:
\begin{verbatim}
    local person = { firstname = "Lewis", lastname = "Reed" }
    person.middlename = "Allan"
    person.firstname = 1942              -- compile-time error
    person.lastname = 2013               -- compile-time error
    print("bye bye " .. person.firstname)
\end{verbatim}

@@

Lua does not set policies on how programmers should define modules,
but it provides mechanisms for organizing a program in modules.
To load a module, Lua first checks whether the module is already
loaded or not.
When the module is not loaded, Lua executes its soure file as a
function, and the value returned by this function is the module.
Thus, Lua modules can be any Lua value.
However, our survey from Section \ref{sec:statistics} confirms that
the convention among Lua programmers is to use tables for defining modules. 
In this case, the fields of the table are the functions and other
values that the module exports.

The simplest way for defining a module in Lua is to declare an
empty table at the begin of the source file,
populate this table with the members that the module should export,
and return this table at the end of the source file. 
The following example shows how Typed Lua handles this case:
\begin{verbatim}
    local mymath = {}
    local RADIANS_PER_DEGREE = 3.14 / 180.0
    mymath.deg = function (x:number)
      return x / RADIANS_PER_DEGREE
    end
    mymath.rad = function (x:number)
      return x * RADIANS_PER_DEGREE
    end
    mymath.pow = function (x:number, y:number)
      return x ^ y
    end
    return mymath
\end{verbatim}

In this example, Typed Lua uses the refinement of table types to
allow the programmer to incrementally build the type of the
module \texttt{mymath}.
Like in Lua, the variable \texttt{RADIANS\string_PER\string_DEGREE}
is private because we declared it as local to the module. 
The \texttt{return} at the end of the source file gives the
final type of the module \texttt{mymath}:
\begin{align*}
\{ & ``deg":\Number \times \Value{*} \rightarrow \Number \times \Nil{*},\\
& ``rad":\Number \times \Value{*} \rightarrow \Number \times \Nil{*},\\
& ``pow":\Number \times \Number \times \Value{*} \rightarrow \Number \times \Nil{*}\}
\end{align*}

Another way of defining modules in Lua is to declare only locals
in the source file and return a table constructor that includes
the members that the module should export at the end of the source file.
The following example shows how Typed Lua handles this case:
\begin{verbatim}
    local RADIANS_PER_DEGREE = 3.14 / 180.0
    local function deg (x:number)
      return x / RADIANS_PER_DEGREE
    end
    local function rad (x:number)
      return x * RADIANS_PER_DEGREE
    end
    local function pow (x:number, y:number)
      return x ^ y
    end
    return {
      deg = deg,
      rad = rad,
      pow = pow,
    }
\end{verbatim}

In this example, the \texttt{return} at the end of the source file
also gives the final type of the module \texttt{mymath}, which is
the same of the previous example.

There is still a third way of defining modules in Lua through the
\texttt{module} function, which has been deprecated.
In this style, the first line of the module calls \texttt{module}.
This function installs a fresh table as the global environment for
the rest of the module, so any assignments to global variables are
field assignments to this table.
The \texttt{module} function also sets an \texttt{\string_M} field
in this table as a circular reference to the table itself,
so the module can end with \texttt{return \string_M},
but this explicit return is not necessary.

Even though this style has been deprecated, our survey from
Section \ref{sec:statistics} indicated that around a third of the
Lua modules that are in LuaRocks still use this style.
Thus, Typed Lua also supports this style:
\begin{verbatim}
    module "mymath"
    local RADIANS_PER_DEGREE = 3.14 / 180.0
    deg = function (x:number)
      return x / RADIANS_PER_DEGREE
    end
    rad = function (x:number)
      return x * RADIANS_PER_DEGREE
    end
    pow = function (x:number, y:number)
      return x ^ y
    end
\end{verbatim}

In this example, Typed Lua treats accesses to global variables as
field accesses to an open table in the top-level scope,
and at the end of file this open table gives the final type of the
module \texttt{mymath}, which is also the same of the previous two
examples.

After we define the module \texttt{mymath}, regardless of adopted style,
users can use it in the standard way:
\begin{verbatim}
    local m = require "mymath"
    print(m.pow(2, 3))                   -- 8
    print(m.pow(2, "foo"))               -- compile-time error
\end{verbatim}

And now the users of the module \texttt{mymath} have the policies that
Typed Lua provides.
The call to \texttt{require} assigns the type of the module \texttt{mymath}
to \texttt{m}, so Typed Lua can catch misuses of the module.

The way Typed Lua handle modules using table types is also
relevant for supporting object-oriented programming, as we discuss in
the next section.
